### 2012-10-25 Docs/Thesis/system_design_and_interface.tex
Updates to system design chapter, still need more combinator descriptions

The representation of signal functions is also discussed in
Section~\ref{section:Implementation-Signal_Functions}. Here it suffices to say
that the use of GADTs permits the construction of values which make use of
uninhabited types as instantiations of type parameters.

The type synonyms {\tt :~>} and {\tt :\textasciicircum:} are included for readability and are
not crucial to the FRP system.

@@

\begin{figure}
\begin{code}
identity :: SF NonInitialized sv sv
constant :: a -> SF NonInitialized SVEmpty (SVSignal a)
never    :: SF NonInitialized SVEmpty (SVEvent a)
asap     :: a -> SF NonInitialized SVEmpty (SVEvent a)
after    :: Double -> a -> SF NonInitialized SVEmpty (SVEvent a)
\end{code}
\hrule
\caption{Basic signal functions.}
\label{figure:basic_signal_functions}
\end{figure}

@@

\begin{figure}
\begin{code}
pureSignal ::    (a -> b) 
              -> SF NonInitialized (SVSignal a) (SVSignal b)
pureEvent  ::    (a -> b)
              -> SF NonInitialized (SVSignal a) (SVSignal b)
\end{code}
\hrule
\caption{Lifting pure functions.}
\label{figure:lifting_pure_functions}
\end{figure}

@@

\begin{figure}
\begin{code}
(>>>)       ::    SF NonInitialized svIn svBetween
               -> SF NonInitialized svBetween svOut
               -> SF NonInitialized svIn svOut
first       ::    SF NonInitialized svIn svOut
               -> SF NonInitialized (SVAppend svIn sv)
                                    (SVAppend svOut sv)
second      ::    SF NonInitialized svIn svOut
               -> SF NonInitialized (SVAppend sv svIn)
                                    (SVAppend sv svOut)
swap        ::    SF NonInitialized (SVAppend svLeft svRight) 
                                    (SVAppend svRight svLeft)
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

@@

The evaluation interface provides a modified state monad which holds a signal
function, together with some additional information, as its state. Rather than
monadic instructions to put and get the state, the monad provides instructions
to trigger an input event, update an input signal, and trigger sampling of
signals in the signal function. Additional state includes the current set of
modifications to the input signals (since the last sample) and a set of
handlers which actuate effects based on output events or changes to the output
signal. The interface combinators are shown in Figure \ref{figure:evaluation_combinators}.

\begin{figure}
\begin{code}
data SVSample sv

data SVEventInput sv
instance SVRoutable SVEventInput sv

data SVSignalUpdate sv
instance SVRoutable SVSignalUpdate sv

data SVHandler out sv

svLeft          :: SVRoutable r =>
                   r svLeft -> r (svLeft :^: svRight)
svRight         :: SVRoutable r =>
                   r svRight -> r (svLeft :^: svRight)

svOcc           :: a -> SVEventInput (SVEvent a)
svSig           :: a -> SVSignalUpdate (SVSignal a)

emptyHandler    :: SVHandler out SVEmpty
signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)
eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)
combineHandlers ::    SVHandler out svLeft
                   -> SVHandler out svRight
                   -> SVHandler out (svLeft :^: svRight)

sample          :: a -> SVSample (SVSignal a)
sampleEvt       :: SVSample (SVEvent a)
sampleNothing   :: SVSample SVEmpty
combineSamples  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (svLeft :^: svRight)

data SFEvalState m svIn svOut
initSFEval ::    SVHandler (m ()) svOut
              -> SVSample svIn
              -> Double
              -> (svIn :~> svOut)
              -> SFEvalState m svIn svOut

data SFEvalT svIn svOut m a
instance MonadTrans (SFEvalT svIn svOut)
instance (Monad m) => Monad (SFEvalT svIn svOut m)
instance (Functor m) => Functor (SFEvalT svIn svOut m)
instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)

runSFEvalT ::    SFEvalT svIn svOut m a
              -> SFEvalState m svIn svOut
              -> m (a, SFEvalState m svIn svOut)

push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()
step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
\end{code}
\hrule
\caption{Evaluation combinators.}
\label{figure:evaluation_combinators}
\end{figure}

\subsection{Initialization and Running}
To evaluate a signal function, one first constructs an initial state
with the {\tt initSFEval} function. This function takes a set of
handlers, a set of initial samples for the signal inputs, the original
time (as a {\tt Double}) and the signal function to evaluate.

The signal function replaces itself after each event occurrence or time
step. The current signal function is maintained in the state. Calling
{\tt push} causes the signal function to react to the supplied event 
occurrence, and events which subsequently occur on the signal function's
output are passed to the supplied handlers. Calling {\tt update} updates
a set of new signal inputs. This set is nullified after it is passed in
along with a new time by calling {\tt step}. This allows various
components of a system to update individual signal inputs in a manner which is
decoupled from the sample timer.

When a monadic evaluation action is passed to the {\tt runSFEvalT} function,
along with a state, the result is an action in the underlying monad. This
action, when run in the underlying monad ({\tt IO}, for example) produces
a pair of an output value and a new state. This new state is what should be 
supplied to the next invocation of {\tt runSFEvalT}.

\subsection{Monad Lifting}

{\tt SFEvalT} is not only a monad, but a monad transformer. This means that
it is parameterized by another monad, and that "running" an {\tt SFEvalT}
action produces an action in this underlying monad. It means also that actions
from this underlying monad may be ``lifted'' into actions of {\tt SFEvalT}.
So if the {\tt IO} monad is the underlying monad, {\tt IO} actions can be 
used to obtain input and arrange for timing. The underlying monad is also the
monad whose actions the output handlers produce, which provides the means of
actuating the output of the signal function.

>>>>

Signal functions with signal vectors as input and output types form a
Haskell {\tt GArrow}~\cite{Megacz2011}. Specifically, the signal function
type constructor (with the initialization parameter fixed) forms the arrow
type, the {\tt SVAppend} type constructor forms the product type, and the
{\tt SVEmpty} type constructor forms the unit type.

The representation of signal functions is discussed in
Section~\ref{section:Implementation-Signal_Functions}. Here it suffices to say
that the use of GADTs permits the construction of values which make use of
uninhabited types as instantiations of type parameters.

The type synonyms {\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for readability and are
not crucial to the FRP system.

@@

\begin{figure}
\begin{code}
-- Pass the input unmodified to the output
identity :: sv :~> sv

-- Produce a signal which is at all times the supplied value
constant :: a -> SVEmpty :~> SVSignal a

-- An event with no occurrences
never    :: SVEmpty :~> SVEvent a

-- An event with one occurrence, as soon as possible after
-- the signal function is initialized
asap     :: a -> SVEmpty :~> SVEvent a

-- An event after the specified amount of time has elapsed.
after    :: Double -> a -> SVEmpty :~> SVEvent a
\end{code}
\hrule
\caption{Basic signal functions.}
\label{figure:basic_signal_functions}
\end{figure}

@@

\begin{figure}
\begin{code}
-- Apply the given function to a signal at all points in time
pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b

-- Apply the given function to each event occurrence
pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
\end{code}
\hrule
\caption{Lifting pure functions.}
\label{figure:lifting_pure_functions}
\end{figure}

@@

The routing combinators are used to combine signal functions, and
to re-arrange signal vectors in order to connect signal functions.
The routing combinators are shown in Figure~\ref{figure:routing_combinators}.

Only those combinators which modify or combine signal functions
({\tt (>>>)}, {\tt first}, {\tt second}) are reactive, and then
only if they inherit their reactivity from the signal function(s)
they modify. The rest do not react to or modify the input in any
way, except to re-arrange it, copy it, or discard it altogether.

\begin{figure}
\begin{code}
-- Use the output of one signal function as the input for another
(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut

-- Pass through the right side of the input unchanged
first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)

-- Pass through the left side of the input unchanged
second :: (svIn :~> svOut) -> (sv :^: svIn) :~> (sv :^: svOut)

-- Swap the left and right sides
swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)

-- Duplicate the input
copy :: sv :~> (sv :^: sv)

-- Ignore the input
ignore :: sv :~> svEmpty

-- Remove an empty vector on the left
cancelLeft :: (SVEmpty :^: sv) :~> sv

-- Add an empty vector on the left
uncancelLeft :: sv :~> (SVEmpty :^: sv)

-- Remove an empty vector on the right
cancelRight :: (sv :^: SVEmpty) :~> sv

-- Add an empty vector on the right
uncancelRight :: sv :~> (sv :^: SVEmpty)

-- Make right-associative
associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))

-- Make left-associative
unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

\subsection{Reactivity}
\label{subsection:System_Design_and_Interface-Combinators-Reactivity}

Reactivity is introduced by means of the {\tt switch} combinator
(Figure~\ref{figure:switch_combinator}). The design of this combinator
allows modular packaging of reactivity. A signal function can determine
autonomously when to replace itself, based only on its input and state,
by emitting an event occurrence carrying its replacement. The combinator
consumes and hides the event carrying the replacement signal function, 
so the reactivity is not exposed by the resulting reactive signal function.

\begin{figure}
\begin{code}
switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
          -> svIn :~> svOut
\end{code}
\hrule
\caption{Combinator for reactivity.}
\label{figure:switch_combinator}
\end{figure}

There are other formulations of a reactive combinator which may be implemented
using the one supplied. These are shown in Figure~\ref{figure:alternate_switching_combinators}
and may be provided in a future version of the TimeFlies library.

\begin{figure}
\begin{code}
-- Alternate version of switch,
-- implemented in terms of supplied version
switch_gen ::    (svIn :~> (svOut :^: SVEvent a))
              -> (a -> svIn :~> svOut)
              -> svIn :~> svOut
switch_gen sf f =
  switch (sf >>> second (pureEventTransformer f))

-- Supplied version in terms of alternate version
switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
          -> svIn :~> svOut
switch sf = switch_gen sf id

-- Repeated switch, which takes replacement signal functions
-- externally.
rswitch ::    (svIn :~> svOut)
           -> (svIn :^: SVEvent (svIn :~> svOut)) :~> svOut
rswitch sf =
  switch (first sf >>> second (pureEventTransformer rswitch))
\end{code}
\hrule
\caption{Alternate combinators for reactivity.}
\label{figure:alternate_switching_combinators}
\end{figure}

@@

The evaluation interface provides a modified state monad which holds a signal
function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
Rather than monadic instructions to put and get the state, the monad provides instructions
to trigger an input event, update an input signal, and trigger sampling of
signals in the signal function. Additional state includes the current set of
modifications to the input signals (since the last sample) and a set of
handlers which actuate effects based on output events or changes to the output
signal.

\begin{figure}
\begin{code}
-- A vector of handlers for outputs
data SVHandler out sv

-- A dummy handler for an empty output
emptyHandler    :: SVHandler out SVEmpty

-- A handler for an updated signal sample
signalHandler   :: (a -> out) -> SVHandler out (SVSignal a)

-- A handler for an event occurrence
eventHandler    :: (a -> out) -> SVHandler out (SVEvent a)

-- Combine handlers for a vector
combineHandlers ::    SVHandler out svLeft
                   -> SVHandler out svRight
                   -> SVHandler out (svLeft :^: svRight)

-- The state maintained when evaluating a signal function
data SFEvalState m svIn svOut

-- Create the initial state for evaluating a signal function
initSFEval ::    SVHandler (m ()) svOut
              -> SVSample svIn
              -> Double
              -> (svIn :~> svOut)
              -> SFEvalState m svIn svOut
\end{code}
\hrule
\caption{State maintained when evaluating a signal function}
\label{figure:evaluation_state}
\end{figure}

In order to initialize the state, the user must supply a set of handlers, the
signal function to evaluate, and initial values for all of the signal inputs
(Figure~\ref{figure:initial_input}).

\begin{figure}
\begin{code}
-- A sample for all leaves of a signal vector
data SVSample sv

-- Create a sample for a signal leaf
sample          :: a -> SVSample (SVSignal a)

-- A dummy sample for an event leaf
sampleEvt       :: SVSample (SVEvent a)

-- A dummy sample for an empty leaf
sampleNothing   :: SVSample SVEmpty

-- Combine two samples
combineSamples  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (svLeft :^: svRight)
\end{code}
\hrule
\caption{Data type for initial input}
\label{figure:initial_input}
\end{figure}

This state can then be passed to a monadic action which will supply input to
the signal function. Inputs are constructed using a simple interface with
functions to construct sample updates and event occurrences, and to specify
their place in the vector (Figure~\ref{figure:ongoing_input}).

\begin{figure}
\begin{code}
-- Class to overload left and right functions
class SVRoutable r where
  svLeft          :: r svLeft -> r (svLeft :^: svRight)
  svRight         :: r svRight -> r (svLeft :^: svRight)

-- An input event occurrence
data SVEventInput sv
instance SVRoutable SVEventInput sv

-- An updated sample for a signal
data SVSignalUpdate sv
instance SVRoutable SVSignalUpdate sv

-- Create an event occurrence
svOcc           :: a -> SVEventInput (SVEvent a)

-- Create an updated sample
svSig           :: a -> SVSignalUpdate (SVSignal a)
\end{code}
\hrule
\caption{Data types for ongoing input.}
\label{figure:ongoing_input}
\end{figure}

The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
parameterized over an underlying monad whose actions may be lifted to
{\tt SFEvalT}. In the usual case, this will be the {\tt IO} monad.

{\tt SFEvalT} actions are constructed using combinators to push events,
update inputs, and step time, as well as actions lifted from the underlying
monad (used to obtain these inputs). An action in the underlying monad
which produces the result and a new state is obtained with the {\tt runSFEvalT}
function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.

\begin{figure}
\begin{code}
-- The evaluation monad
data SFEvalT svIn svOut m a
instance MonadTrans (SFEvalT svIn svOut)
instance (Monad m) => Monad (SFEvalT svIn svOut m)
instance (Functor m) => Functor (SFEvalT svIn svOut m)
instance (Monad m, Functor m) => Applicative (SFEvalT svIn svOut m)
instance (MonadIO m) => MonadIO (SVEvalT svIn svOut m)

-- Obtain an action in the underlying monad
-- from an SFEvalT and a new state.
runSFEvalT ::    SFEvalT svIn svOut m a
              -> SFEvalState m svIn svOut
              -> m (a, SFEvalState m svIn svOut)

-- Push an event occurrence.
push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()

-- Update the value of an input signal sample
-- (not immediately observed)
update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()

-- Step forward in time, observing the updated signal values
step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
\end{code}
\hrule
\caption{Evaluation combinators}
\label{figure:evaluation_combinators}
\end{figure}
